An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullets

### 2021-10-07 CMPSpaper_writing/CMPSpaper_writing.Rmd

---
title: An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullets
header-includes:
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{longtable}
  - \usepackage{adjustbox}            
  - \usepackage[font=small,skip=5pt]{caption}
  - \usepackage{subcaption}
  - \usepackage{afterpage}
  - \usepackage{bbm}
author:
  # see ?rjournal_article for more information
  - name: Wangqian W. Ju
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
    url: https://github.com/willju-wangqian
    email:  wju@iastate.edu
    orcid: 0000-0002-9977-377X
  - name: Heike Hofmann
    url: https://github.com/heike
    email: hofmann@iastate.edu
    orcid: 0000-0002-9079-593X
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
thanks: |
  This work was partially funded by the Center for Statistics and Applications in Forensic Evidence (CSAFE) through Cooperative Agreement 70NANB20H019 between NIST and Iowa State University, which includes activities carried out at Carnegie Mellon University, Duke University, University of California Irvine, University of Virginia, West Virginia University, University of Pennsylvania, Swarthmore College and University of Nebraska, Lincoln.    
abstract: >
  In this paper, we introduce the R package CMPS, an open-source implementation of the Congruent Matching Profile Segments (CMPS) method developed at the National Institute of Standards and Technology (NIST) for objective comparison of striated tool marks. The functionality of the package is showcased by examples of bullet signatures that come with the package. Graphing tools are implemented in the package as well for users to assess and understand the CMPS results. Initial tests were performed on bullet signatures generated from two sets of 3D scans in the Hamby study under the framework suggested by the R package bulletxtrctr. New metrics based on CMPS scores are introduced and compared with existing metrics. A measure called Sum of Squares Ratio is included, and how it can be used for evaluating different scans, metrics, or parameters is showcased with the Hamby study data sets. An open-source implementation of the CMPS algorithm makes the algorithm more accessible, generates reproducible results, and facilitates further studies of the algorithm such as methods comparison.
preamble: |
  % Any extra LaTeX you need in the preamble

@@

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "\\textwidth",
  fig.align = "center",
#  cache=TRUE,
  dpi = 100
)
```

@@

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method.
@cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations.
Although @cmps conceptually described the CMPS algorithm in their paper, the authors did not release an implementation of their method.
Thus, our effort here is to introduce the CMPS method to the `R` [@R] community and provide an open-source, publicly available implementation of the algorithm to use, review, and improve.
Our implementation is made available as part of the `R` package `cmpsR` on GitHub at \url{https://github.com/willju-wangqian/cmpsR}.

@@

Most algorithms for comparing striation marks are based on the digitized signatures and produce a similarity score [@song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940].
The congruent matching profile segments (CMPS) algorithm, developed by @cmps for "objective comparison of striated tool marks", is one such algorithm.
The algorithm's main idea is to take a set of consecutive and non-overlapping basis segments from the comparison and for each segment find the "best" registration position on the reference (the other bullet signature) with respect to their cross-correlation values.
From a comparison of these registration positions, a **congruent registration position** is identified, and the number of basis segments taking the congruent registration position is the CMPS score.
Note that researchers in @cmps refers to the origin of basis segments as the reference, but in this paper we refer to it as the comparison.
High CMPS scores are achieved between more similar signatures and are therefore indicative of a same-source pair.
Low scores between pairs of signatures are attributed to different source pairs.
However, a specific threshold of the CMPS score to distinguish between same-source and different-source comparisons is not provided in @cmps, instead the threshold depends on the underlying structure of the data and the choice of parameters. \hh{XXX  In a legal setting this variability is problematic because it allows for situations in which experts could choose parameters based on a wished-for outcome. }
Further research is needed to understand how to determine optimal threshold settings.
The CMPS algorithm can assist firearm examiners with drawing a conclusion about the source of a comparison pair.
Unfortunately, @cmps did not release any code or specific parameter settings for their implementation of the CMPS algorithm.
Thus, in this paper we present an open-source implementation of the CMPS algorithm in the R package cmpsR available from Github \url{https://github.com/willju-wangqian/cmpsR}. This publicly available implementation calculates the CMPS score of a comparison using the following code:

@@

The section "Implementation" will go through the algorithm and show how to use the \pkg{cmpsR} package. 
A further example that illustrates the main points are also included.
The section "Results" presents the results of evaluating the `cmpsR` package using Hamby set 252 and Hamby set 44. The results from Hamby set 252 are used to verify that our implementation is, at least qualitatively, comparable to the algorithm described in @nist. Results from Hamby 44 show the need for a further investigation of the parameter choices even in the case of bullets fired from the same barrels. \hh{We suggest a sum of square ratio ...}
The last section covers some final discussion and conclusions.

@@

In the remainder of the paper we showcase the use of the CMPS functionality on some examples and present the results of applying it to two datasets.

@@

The `cmpsR` package is not publicly available from CRAN yet, and its development version is available from Github and can be installed by

```{r, eval=FALSE}
# install.packages("remotes") 
remotes::install_github("willju-wangqian/CMPS")
```

@@

The `cmpsR` package contains a simple example to illustrate the basic usage of the package.
The data in this example are twelve bullet signatures obtained from two bullets in Hamby set 252 [@hamby].
The procedure for generating signatures from high-resolution 3D topographic scans of bullet lands used here follows the methodology described in @aoas (as discussed above).
The two bullets under consideration are known to have been fired from the same gun barrel, so for the 36 pairwise land-by-land comparisons, six comparisons are from same-source pairs (known matches) while thirty are from different-source pairs (known non-matches).
To access the example data, we use

@@

We investigated the effects of different sets of parameters at the example of both Hamby sets 252 and 44.
More specifically, we investigated the separation achieved using the CMPS implementation under various segment lengths (controlled by the parameter   `seg_length`). Specifically, we fixed the parameter `npeaks.set` that controls the number of peaks at each segment level to be `npeaks.set = c(5,3,1)` and modified the parameter `seg_length`. 
\autoref{fig:param_seg_plot} shows that the default values of `seg_length`  for Hamby 252 and Hamby 44 (50 for Hamby 252 and 61 for Hamby 44, which represent 78.125 \textmu m and 78.69 \textmu m, respectively) result in high values of the sum of squares ratio, no matter which CMPS metrics are used for an evaluation. 
However, we also see, that for Hamby set 252 a basis segment length of 75 is a better choice than the default segment length; a basis segment length of 122 results in a higher value of sum of squares ratio for Hamby 44.

\autoref{fig:param_npeak_plot} shows results of the CMPS algorithm using different stategies for identifying peaks in the correlation structure between signatures. Basis segment length is fixed to the default level for this evaluation. 
As can be seen in \autoref{fig:param_npeak_plot}, the default value of `npeaks.set` (`npeaks.set = c(5,3,1)`) leads to promising results in terms of the sum of squares ratio; however, other choices of `npeaks.set` match and  exceed this sum of squares ratio value.

@@

In this paper we present the `cmpsR` package, an open-source implementation of the Congruent Matching Profile Segments (CMPS) algorithm [@cmps], and apply it to two datasets in Hamby study [@hamby] to show its potential for further research.
The CMPS algorithm was proposed by NIST in 2019 and was made for objective tool marks comparisons.
We introduce the basic logic of the CMPS algorithm and lay out its implementation in the `cmpsR` package.
We also showcase the functionality of the `cmpsR` package with a small dataset example that is included in the package.
In the `cmpsR` package we implement some graphing tools for users to visualize results and to gain a better understanding of both the algorithm and the results.

\wj{Additionally, we propose new metrics based on the CMPS scores and compare the new metrics with the existing metrics. 
We also introduce a principled evaluation framework of algorithmic results based on a measure called the sum of squares ratio. 
We showcase the implementation with two datasets in Hamby study (Hamby set 252 and Hamby set 44) and compare the CMPS algorithm using different sets of parameters and different metrics, following the evaluation framework based on the sum of squares ratio. }
<!-- Additionally, we apply the `cmpsR` package to two datasets in Hamby study (Hamby set 252 and Hamby set 44).  -->
The results obtained are promising:  we were both able to reproduce the results in @cmps qualitatively and achieve a clear separation between the known match (KM) comparisons and known non-match (KNM) comparisons in another bullet study.
\wj{However, the comparisons among different sets of parameters suggest that the optimal choice of the parameters varies for different datasets. 
And further research can follow the proposed evaluation framework to have a comprehensive cross-validation on parameters used the algorithm.
Furthermore, we compared our implemented CMPS algorithm with the random forest model proposed by @aoas and suggest adding the CMPS score as an additional feature in the random forest model.}

The `cmpsR` package provides an open-source implementation and introduces the CMPS algorithm to the `R` community.

Open-source implementations not only allow a peer-review but also facilitate further research, such as parameter cross-validations, method comparisons, and the development of statistical methods for modeling KM and KNM CMPS score distributions, which can then be used for error-rate estimations and fair applications that were called for by the PCAST [@pcast] report.

>>>>

---
title: An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullets
header-includes:
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{longtable}
  - \usepackage{adjustbox}            
  - \usepackage[font=small,skip=5pt]{caption}
  - \usepackage{subcaption}
  - \usepackage{afterpage}
  - \usepackage{bbm}
author:
  # see ?rjournal_article for more information
  - name: Wangqian W. Ju
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
    url: https://github.com/willju-wangqian
    email:  wju@iastate.edu
    orcid: 0000-0002-9977-377X
  - name: Heike Hofmann
    url: https://github.com/heike
    email: hofmann@iastate.edu
    orcid: 0000-0002-9079-593X
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
thanks: |
  This work was partially funded by the Center for Statistics and Applications in Forensic Evidence (CSAFE) through Cooperative Agreement 70NANB20H019 between NIST and Iowa State University, which includes activities carried out at Carnegie Mellon University, Duke University, University of California Irvine, University of Virginia, West Virginia University, University of Pennsylvania, Swarthmore College and University of Nebraska, Lincoln.    
abstract: >
  In this paper, we introduce the R package cmpsR, an open-source implementation of the Congruent Matching Profile Segments (CMPS) method developed at the National Institute of Standards and Technology (NIST) for objective comparison of striated tool marks. The functionality of the package is showcased by examples of bullet signatures that come with the package. Graphing tools are implemented in the package as well for users to assess and understand the CMPS results. Initial tests were performed on bullet signatures generated from two sets of 3D scans in the Hamby study under the framework suggested by the R package bulletxtrctr. New metrics based on CMPS scores are introduced and compared with existing metrics. A measure called Sum of Squares Ratio is included, and how it can be used for evaluating different scans, metrics, or parameters is showcased with the Hamby study data sets. An open-source implementation of the CMPS algorithm makes the algorithm more accessible, generates reproducible results, and facilitates further studies of the algorithm such as methods comparison.
preamble: |
  % Any extra LaTeX you need in the preamble

@@

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "\\textwidth",
  fig.align = "center",
#  cache=TRUE,
  dpi = 100,
  warning = FALSE
)
```

@@

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method.
@cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations.
Although @cmps conceptually described the CMPS algorithm in their paper, the authors did not release an implementation of their method.
Thus, our effort here is to introduce the CMPS method to the `R` [@R] community and provide an open-source, publicly available implementation of the algorithm to use, review, and improve.
Our implementation is made available as part of the `R` package \pkg{cmpsR} on GitHub at \url{https://github.com/willju-wangqian/cmpsR}.

@@

Most algorithms for comparing striation marks are based on the digitized signatures and produce a similarity score [@song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940].
The congruent matching profile segments (CMPS) algorithm, developed by @cmps for "objective comparison of striated tool marks", is one such algorithm.
The algorithm's main idea is to take a set of consecutive and non-overlapping basis segments from the comparison and for each segment find the "best" registration position on the reference (the other bullet signature) with respect to their cross-correlation values.
From a comparison of these registration positions, a **congruent registration position** is identified, and the number of basis segments taking the congruent registration position is the CMPS score.
Note that researchers in @cmps refers to the origin of basis segments as the reference, but in this paper we refer to it as the comparison.
High CMPS scores are achieved between more similar signatures and are therefore indicative of a same-source pair.
Low scores between pairs of signatures are attributed to different source pairs.
However, a specific threshold of the CMPS score to distinguish between same-source and different-source comparisons is not provided in @cmps, instead the threshold depends on the underlying structure of the data and the choice of parameters. \hh{XXX  In a legal setting this variability is problematic because it allows for situations in which experts could choose parameters based on a wished-for outcome. }
Further research is needed to understand how to determine optimal threshold settings.
The CMPS algorithm can assist firearm examiners with drawing a conclusion about the source of a comparison pair.
Unfortunately, @cmps did not release any code or specific parameter settings for their implementation of the CMPS algorithm.
Thus, in this paper we present an open-source implementation of the CMPS algorithm in the R package \pkg{cmpsR} available from Github \url{https://github.com/willju-wangqian/cmpsR}. This publicly available implementation calculates the CMPS score of a comparison using the following code:

@@

The section "Implementation" will go through the algorithm and show how to use the \pkg{cmpsR} package. 
A further example that illustrates the main points is also included.
The section "Results" presents the results of evaluating the \pkg{cmpsR} package using Hamby set 252 and Hamby set 44. The results from Hamby set 252 are used to verify that our implementation is, at least qualitatively, comparable to the algorithm described in @cmps. Results from Hamby 44 show the need for a further investigation of the parameter choices even in the case of bullets fired from the same barrels. \wj{We suggest new CMPS metrics that summarize land-level CMPS scores and a sum of squares ratio that can be used to evaluate algorithmic results.}
<!-- \hh{We suggest a sum of square ratio ...} -->
The last section covers some final discussion and conclusions.

@@

In the remainder of the paper we showcase the use of the \pkg{cmpsR} functionality on some examples and present the results of applying it to two datasets.

@@

The \pkg{cmpsR} package is not publicly available from CRAN yet, and its development version is available from Github and can be installed by

```{r, eval=FALSE}
# install.packages("remotes") 
remotes::install_github("willju-wangqian/cmpsR")
```

@@

The \pkg{cmpsR} package contains a simple example to illustrate the basic usage of the package.
The data in this example are twelve bullet signatures obtained from two bullets in Hamby set 252 [@hamby].
The procedure for generating signatures from high-resolution 3D topographic scans of bullet lands used here follows the methodology described in @aoas (as discussed above).
The two bullets under consideration are known to have been fired from the same gun barrel, so for the 36 pairwise land-by-land comparisons, six comparisons are from same-source pairs (known matches) while thirty are from different-source pairs (known non-matches).
To access the example data, we use

@@

We investigated the effects of different sets of parameters at the example of both Hamby sets 252 and 44.
More specifically, we investigated the separation achieved using the \pkg{cmpsR} implementation under various segment lengths (controlled by the parameter `seg_length`). Specifically, we fixed the parameter `npeaks.set` that controls the number of peaks at each segment level to be `npeaks.set = c(5,3,1)` and modified the parameter `seg_length`. 
\autoref{fig:param_seg_plot} shows that the default values of `seg_length`  for Hamby 252 and Hamby 44 (50 for Hamby 252 and 61 for Hamby 44, which represent 78.125 \textmu m and 78.69 \textmu m, respectively) result in high values of the sum of squares ratio, no matter which CMPS metrics are used for an evaluation. 
However, we also see, that for Hamby set 252 a basis segment length of 75 is a better choice than the default segment length; a basis segment length of 122 results in a higher value of sum of squares ratio for Hamby 44.

\autoref{fig:param_npeak_plot} shows results of the CMPS algorithm using different strategies for identifying peaks in the correlation structure between signatures. Basis segment length is fixed to the default level for this evaluation. 
As can be seen in \autoref{fig:param_npeak_plot}, the default value of `npeaks.set` (`npeaks.set = c(5,3,1)`) leads to promising results in terms of the sum of squares ratio; however, other choices of `npeaks.set` match and  exceed this sum of squares ratio value.

@@

In this paper we present the \pkg{cmpsR} package, an open-source implementation of the Congruent Matching Profile Segments (CMPS) algorithm [@cmps], and apply it to two datasets in Hamby study [@hamby] to show its potential for further research.
The CMPS algorithm was proposed by NIST in 2019 and was made for objective tool marks comparisons.
We introduce the basic logic of the CMPS algorithm and lay out its implementation in the \pkg{cmpsR} package.
We also showcase the functionality of the \pkg{cmpsR} package with a small dataset example that is included in the package.
In the \pkg{cmpsR} package we implement some graphing tools for users to visualize results and to gain a better understanding of both the algorithm and the results.

\wj{Additionally, we propose new metrics based on the CMPS scores and compare the new metrics with the existing metrics. 
We also introduce a principled evaluation framework of algorithmic results based on a measure called the sum of squares ratio. 
We showcase the implementation with two datasets in Hamby study (Hamby set 252 and Hamby set 44) and compare the CMPS algorithm using different sets of parameters and different metrics, following the evaluation framework based on the sum of squares ratio. }
<!-- Additionally, we apply the \pkg{cmpsR} package to two datasets in Hamby study (Hamby set 252 and Hamby set 44).  -->
The results obtained are promising:  we were both able to reproduce the results in @cmps qualitatively and achieve a clear separation between the known match (KM) comparisons and known non-match (KNM) comparisons in another bullet study.
\wj{However, the comparisons among different sets of parameters suggest that the optimal choice of the parameters varies for different datasets. 
And further research can follow the proposed evaluation framework to have a comprehensive cross-validation on parameters used the algorithm.
Furthermore, we compared our implemented CMPS algorithm with the random forest model proposed by @aoas and suggest adding the CMPS score as an additional feature in the random forest model.}

\wj{The open-source implementation of the CMPS algorithm provided by the \pkg{cmpsR} package is just one step towards the world of open science, which is important to fields like the forensic science where transparency and accuracy are critical to fair justice. 
Open-source implementations not only allow a peer-review but also facilitate further research, such as parameter cross-validations, method comparisons, and the development of statistical methods for modeling KM and KNM CMPS score distributions, which can then be used for error-rate estimations and fair applications that were called for by the PCAST [@pcast] report.
However, having open-source implementations is not enough for open science. 
In order to actually build the world of open science, many other efforts, such as open evaluation, open access publication, open educational resources, etc., are still needed. 
The database built and maintained by the National Institute of Standards and Technology is a good example of open data. 
And what we are aiming for is an open system that is able to collect open results, compare multiple algorithms using multiple datasets, and evaluate algorithmic variation and accuracy. 
The evaluation metric we proposed in this paper can be used to compare different algorithms or even different datasets and is a snippet of this open system. 
But the core of this open system is the open science culture and contributions of the community.}
